Bayesian inference of relative fitness on high-throughput pooled competition assays

### 2023-11-20 doc/_bayes_hierarchical.qmd
remove repeated label

Writing Bayes' theorem for the first term in @eq-hier_bayes_factorized_replicate
results in
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\pi(
    \{\underline{\underline{F}}_{[j]}\} \mid
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
)
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
),
$${#eq-hier_bayes}
where we leave the conditioning on the population mean fitness as we did in
@sec-bayes_mutfit. This expression can be simplified in two ways. First, the
frequency values for each experimental replicate depend directly on the local
fitness values and the corresponding population mean fitness, as the
relationship between experimental replicates only occurs through the relative
fitness values. Therefore, we can write
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
\right]
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
).
$${#eq-hier_bayes_exp_indep}
Second, the relationship between the hyper-fitness and the local fitness values
allows us to write their joint distribution as a conditional distribution where
local fitness values depend on the global hyper-fitness value, obtaining
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
    \pi(
        \underline{s}^M_{[j]} \mid
        \underline{\theta}^M
    )
\right] 
\pi(
    \underline{\theta}^M
).
$${#eq-hier_bayes_exp_indep}
Notice we removed the conditioning on the population mean fitness as our prior
expectations of what the global hyper-fitness or local fitness value might be do
not depend on these nuisance parameters.

The first term on the right-hand side of @eq-hier_bayes_exp_indep takes the same
functional form as the one derived in @sec-bayes_mutfit. Therefore, all we are
left with is to determine the functional forms for the hyper–prior
$\pi(\underline{\theta}^M)$, and the conditional probability
$\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)$. In analogy to the 
assumptions used for the fitness values in @sec-bayes_mutfit, we define the 
value of each hyper-fitness as independent. This means that we have
$$
\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
$${#eq-hier_hyper_prior}
Furthermore, we assume this prior is of the form
$$
\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
$${#eq-hier_hyper_prior_02}
where $\mu_{\theta^{(m)}}$ and $\sigma_{\theta^{(m)}}$ are user-defined 
parameters encoding the prior expectations on the fitness values.

>>>>

Writing Bayes' theorem for the first term in @eq-hier_bayes_factorized_replicate
results in
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\pi(
    \{\underline{\underline{F}}_{[j]}\} \mid
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
)
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
),
$${#eq-hier_bayes}
where we leave the conditioning on the population mean fitness as we did in
@sec-bayes_mutfit. This expression can be simplified in two ways. First, the
frequency values for each experimental replicate depend directly on the local
fitness values and the corresponding population mean fitness, as the
relationship between experimental replicates only occurs through the relative
fitness values. Therefore, we can write
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
\right]
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\bar{s}}_{T[j]}\}
).
$${#eq-hier_bayes_exp_indep}
Second, the relationship between the hyper-fitness and the local fitness values
allows us to write their joint distribution as a conditional distribution where
local fitness values depend on the global hyper-fitness value, obtaining
$$
\pi(
    \underline{\theta}^M, 
    \{\underline{s}^M_{[j]}\} \mid
    \{\underline{\underline{F}}_{[j]}\},
    \{\underline{\bar{s}}_{T[j]}\}
) \propto
\prod_{j=1}^E\left[
    \pi(
        \underline{\underline{F}}_{[j]} \mid
        \underline{s}^M_{[j]},
        \underline{\bar{s}}_{T[j]}
    )
    \pi(
        \underline{s}^M_{[j]} \mid
        \underline{\theta}^M
    )
\right] 
\pi(
    \underline{\theta}^M
).
$${#eq-hier_bayes_exp_indep_02}
Notice we removed the conditioning on the population mean fitness as our prior
expectations of what the global hyper-fitness or local fitness value might be do
not depend on these nuisance parameters.

The first term on the right-hand side of @eq-hier_bayes_exp_indep_02 takes the
same functional form as the one derived in @sec-bayes_mutfit. Therefore, all we
are left with is to determine the functional forms for the hyper–prior
$\pi(\underline{\theta}^M)$, and the conditional probability
$\pi(\underline{s}^M_{[j]} \mid \underline{\theta}^M)$. In analogy to the 
assumptions used for the fitness values in @sec-bayes_mutfit, we define the 
value of each hyper-fitness as independent. This means that we have
$$
\pi(\underline{\theta}^M) = \prod_{m=1}^M \pi(\theta^{(m)}).
$${#eq-hier_hyper_prior}
Furthermore, we assume this prior is of the form
$$
\theta^{(m)} \sim \mathcal{N}(\mu_{\theta^{(m)}}, \sigma_{\theta^{(m)}}),
$${#eq-hier_hyper_prior_02}
where $\mu_{\theta^{(m)}}$ and $\sigma_{\theta^{(m)}}$ are user-defined 
parameters encoding the prior expectations on the fitness values.
